---
title: "Bibliography Setup"
author:
  name: "Thomas Roppelt"
  affiliation: "Friedrich-Alexander-Universität Erlangen-Nürnberg (FAU)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: library.bib
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r readme, child="README.md"}
```

## Bibliography Managers  {.alert .alert-success}
Bibliography managers are important to keep all the literature you work with well organized.
This page will help you understand pros and cons of different bibliography managers as well as help you set up Zotero, a free and open-source bibliography manager.

## Comparison
There are many different bibliography managers available, e.g. Citavi, Colwiz, EndNote, JabRef, Mendeley, Zotero and ZoteroBib.
Since we focus on open-source software, software like Citavi and EndNote are out of question.
Of the named software, only JabRef, Zotero and ZoteroBib are free and open source.
For a detailed feature comparison, you may want to look [here (german)](https://ub.fau.de/schreiben-publizieren/literaturverwaltung/) or [here (english)](https://en.wikipedia.org/wiki/Comparison_of_reference_management_software).
While the following paragraphs will cover Zotero, please feel free to use any of the other open source bibliography managers. As long as they can handle the [BibTex format](https://de.wikipedia.org/wiki/BibTeX) it should be all fine.

## Zotero setup
To get started with Zotero, there are some components that should be set up:
-Zotero itself
  -via chocolatey package manager (recommended)
    Run in bash or command line: `choco install zotero`
  -download and install via the [Zotero website](https://www.zotero.org/)
-the BetterBibTex Addon for Zotero.
  Among other things, this enables an automatic export of the bib file.
  Prerequisite: Zotero.
  You can find the installation instructions [Here](https://retorque.re/zotero-better-bibtex/installation/).
-the Zotero-picker Addon for Atom
  It enables you to quote and cite directly in Atom.
  Prerequisites: Zotero, BetterBibTex Addon
  The installation instructions are [Here](https://atom.io/packages/zotero-picker).


## References
